One-armed spirals in locally isothermal, radially structured self-gravitating discs

### 2015-01-07 paper/appendix.tex
major revision in response to report

>>>>

\section{The confined spiral as an external potential}\label{disc-planet}
Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $M_\mathrm{ring}$ is the disc mass contained within
$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
and $\beta = R/\overline{R}$. This form of
$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
potential of an external satellite on a circular orbit
\citep{goldreich79}.

%For an order-of-magnitude exercise, we simply take 
%\begin{align}
 % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
%\end{align}
%for the potential associated with a disturbance located at
%$R=\overline{R}$, where $M_\mathrm{ring}$

We expect the external potential to exert a torque on the disc at the
Lindblad and co-rotation resonances. At the outer Lindblad resonance
(OLR), this torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 -
      \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2, 
\end{align}   
where a Keplerian disc has been assumed and subscript $L$ denotes
evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
exist for the pattern speeds observed in our simulations.)

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
  &\times
  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
    + 2\left[1 - 
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
\left[b_{1/2}^1(\beta_c)\right]^2    
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

### 2015-01-07 paper/discussion.tex
major revision in response to report

\section{Discussion}\label{discussions}

@@

\subsection{The role of self-gravity and disc structure} 
It is important to note that in the above discussion, we only considered a self-gravitating
perturbation to obtain an appropriate wavenumber $k$ needed to
evaluate Eq. \ref{theoretical_rate}. 
The background torque density $T_\mathrm{BG}$ does not depend on  
self-gravity explicitly, so self-gravity is not directly
responsible for instability.

However, 
our numerical simulations show that the $m=1$ spiral is
confined between two $Q$-barriers, where real solutions to the local
dispersion relation is possible (Eq. \ref{wavenumber}), as a result of the adopted disc
structure. Thus, in our disc models the main role of disc structure and self-gravity is to 
allow a neutral mode to be set up and confined, which is then
destabilised through the background torque density.

In order to confine an $m=1$ spiral between two $Q$-barriers, we
should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
\begin{align}
  \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
\end{align}
Then it may be possible to have two $Q$-barriers when the $Q$ profile
of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
for decreasing (increasing) $R$. One possibility is a gap
opened by a giant planet in a constant-$Q$ disc. In that case $Q$
rises rapidly towards the gap centre since it is a region of low
surface density. The outer gap edge is then a potential site for the
growth of a low-frequency one-armed spiral through the background
torque.

%talk about additional simulations at high Q
%spiral mode as external potential to wider disc (lp11)
%interaction across co-rotation

\subsection{The confined spiral as an external potential}
One possible effect of self-gravity is to allow the one-armed spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}. Here, we estimate the magnitude of this effect
using basic results from disc-planet theory \citep[see, e.g.][and
references therein]{papaloizou07}.

Let us treat the one-armed spiral confined in $R\in[R_1,R_2]$ as an  
external potential of the form $\Phi_\mathrm{ext}(R)\cos{\left(\phi -
    \Omega_pt\right)}$. We take 
\begin{align} 
  \Phi_\mathrm{ext}
  =-\frac{GM_\mathrm{ring}}{\overline{R}}b^{1}_{1/2}(\beta),   
\end{align}
where $M_\mathrm{ring}$ is the disc mass contained within
$R\in[R_1,R_2]$, $\overline{R} = (R_1+R_2)/2$ is the approximate radial
location of the spiral, $b_{n}^m(\beta)$ is the Laplace coefficient
and $\beta = R/\overline{R}$. This form of
$\Phi_\mathrm{ext}$ is the $m=1$ component of the gravitational
potential of an external satellite on a circular orbit
\citep{goldreich79}.

%For an order-of-magnitude exercise, we simply take 
%\begin{align}
 % \Phi_\mathrm{ext} = -\frac{GM_\mathrm{ring}}{|R - \overline{R}|}, 
%\end{align}
%for the potential associated with a disturbance located at
%$R=\overline{R}$, where $M_\mathrm{ring}$

We expect the external potential to exert a torque on the disc at the
Lindblad and co-rotation resonances. At the outer Lindblad resonance
(OLR), this torque is 
\begin{align}
  \Gamma_L =
  \frac{\pi^2\Sigma_L}{3\Omega_L\Omega_p}
  \left[\left.R_L\frac{d\Phi_\mathrm{ext}}{dR}\right|_L + 2\left(1 -
      \frac{\Omega_p}{\Omega_L}\right)\Phi_\mathrm{ext}\right]^2, 
\end{align}   
where a Keplerian disc has been assumed and subscript $L$ denotes
evaluation at the OLR, $R=R_L$. (The inner Lindblad resonance does not
exist for the pattern speeds observed in our simulations.)

If we associate the external potential with angular momentum 
$J_\mathrm{ext}  = M_\mathrm{ring}\overline{R}^2\Omega_p$, we can
calculate a growth rate $\gamma_L=\Gamma_L/J_\mathrm{ext}$. Then 
\begin{align}
  \frac{\gamma_L}{\Omega_p} = &\frac{\pi
    h}{3Q_L}\left(\frac{M_p}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)\left(\frac{R_c}{\overline{R}}\right) 
  ^3\left(\frac{R_L}{R_c}\right)^{-3/2}\notag\\ 
  &\times
  \left\{\frac{R_L}{\overline{R}}\left.\frac{db_{1/2}^1}{d\beta}\right|_L
    + 2\left[1 - 
      \left(\frac{R_c}{R_L}\right)^{-3/2}\right]b_{1/2}^1(\beta_L)\right\}^2. 
\end{align}
%\begin{align}
 % \frac{\gamma_L}{\Omega_p} \sim &\frac{\pi h}{3
  %  Q_L}\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_L}{\overline{R}}\right)^{-1/2}\left(\frac{R_c}{\overline{R}}\right)^{5/2}\notag\\
 % &\times \left\{ \frac{R_L}{R_L - \overline{R}} - 2\left[1 -
  %    \left(\frac{R_c}{R_L}\right)^{-3/2}\right]\right\}^2\frac{R_c^2}{\left(R_L
   % - \overline{R}\right)^2}. 
%\end{align}
Inserting $h=0.05$, $Q_L=10$, $M_\mathrm{ring} = 0.05M_*$,
$R_L=7.2R_0$, $R_c=4.4R_0$ and $\overline{R}=1.5R_0$ from our fiducial
FARGO simulation, we get
\begin{align}
  \gamma_L \sim 5\times10^{-4}\Omega_p. 
\end{align}
%\begin{align}
 % \gamma_L \sim 0.012\Omega_p. 
%\end{align}
%This represents an upper limit on the effect of the OLR on the spiral
%disturbance, and is much smaller than the growth rate measured in the
%simulation ($\gamma\sim0.1\Omega_p$). We conclude that for our models,
%the OLR has negligible effect on the one-armed spiral.  
For the co-rotation torque, we use the result
\begin{align}
  \Gamma_c = \left.
    \pi^2\Phi_\mathrm{ext}^2\left(\frac{d\Omega}{dR}\right)^{-1}\frac{d}{dR}\left(\frac{2\Sigma}{\Omega}\right)\right|_{c}     
\end{align}
for a Keplerian disc, where subscript $c$ denotes evaluation at
co-rotation radius $R=R_c$. For a power-law surface density profile
$\Sigma\propto R^{-s}$ we have
\begin{align}
\frac{\gamma_c}{\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_c}
\left(\frac{M_\mathrm{ring}}{M_*}\right)\left(\frac{R_c}{\overline{R}}\right)^4\left(s-\frac{3}{2}\right) 
\left[b_{1/2}^1(\beta_c)\right]^2    
\end{align}
%\begin{align}
%  \frac{\Gamma_C}{J_\mathrm{ext}\Omega_p} = \frac{4}{3}\frac{\pi h}{Q_C} \left(s -
 %   \frac{3}{2}\right)\left(\frac{M_\mathrm{ring}}{M_*}\right)\frac{R_c^4}{\overline{R}^2\left(R_c
 %     - \overline{R}\right)^2}.   
%\end{align}
Using the above parameter values with $s=2$ and $Q_c=10$, we obtain a
rate
\begin{align}
  \gamma_c\sim 6\times 10^{-4}\Omega_p. 
\end{align}

These estimates are much smaller than that due to
the imposed temperature gradient as measured in the simulations
($\gamma\sim 0.1\Omega_p$). We conclude that for our disc models, the
Lindblad and co-rotation resonances have negligible effects on the
growth of the one-armed spiral in the inner disc. We confirmed this with
additional FARGO simulations with a reduced radial domain size,
setting $R_\mathrm{max}=3R_0$, which still developed the one-armed
spiral.

The torque exerted on the disc at the OLR by an
external potential is positive, while that at co-rotation 
depends on the gradient of potential vorticity there
\citep{goldreich79}. For our disc models with surface density
$\Sigma\propto R^{-2}$ in the outer disc, this co-rotation torque is
positive. Interpreting the one-armed spiral as an external potential
for the outer disc, this means that the spiral loses
angular momentum by launching density waves with positive angular momentum at
the OLR, and by applying a positive co-rotation torque on the disc. 
In principle, this interaction is destabilising because the
one-armed spiral has negative angular momentum
\citep{lin11b}. This effect is small in our simulations, but it  
could be significant in other parameter regimes. %Furthermore, this
%non-local interaction may be important for the long-term evolution of
%the outer disc, since it gains angular momentum from the

{\bf
  \subsection{Relevance to protoplanetary discs}

>>>>

\section{Discussion}\label{discussions} 
   % specific to our numerical 
   % simulations%  and may or may not be relevant to generic low-frequency
   %   % modes destabilised by a forced temperature gradient.  
   %

{\bf
  \subsection{Applicability to circumstellar discs}

  \subsubsection{Thermodynamic requirements}

  % talk about additional simulations at high Q

  %issues with fragmentation 
 % requirement for fast cooling  
  % fast cooling fragments 
  % stabilised by irradiation
  % background torque and fragmentation may not be mutually exclusive
  % (speculations) 
  % these studies do no account for possible global disk structure 
  %Clarke (2009), Rafikov (2005), and Rice & Armitage (2009)
  %no fragmentation in our case because disc is maintained at slightly
  %higher values of Q. no fragmentation of one arm spiral because
  %co-rotation radius isn't on it 
  %if irradiation 
  %one-armed low frequency spiral probably will not fragment because
  %co-rotation is far away

  A locally isothermal equation of state represents the limit
  of infinitely short cooling (and heating) timescales, so the
  disc temperature instantly returns to its initial value when
  perturbed.  A short cooling timescale $t_c$ can occur in the outer
  parts of protoplanetary discs 
  \citep{rafikov05,clarke09,rice09,cossins10}.  %rafikov07 (convection)
  However, if a disc with 
  $Q\simeq 1$ is cooled (towards zero temperature) on a timescale
  $t_c\ll\Omega_k^{-1}$, it will fragment
  \citep{gammie01,paarkedooper12}.

  %internal heating?
  To avoid fragmentation, the disc must be heated to
  maintain $Q$ above the threshold for fragmentation \citep[about
  $Q\lesssim 1.4$ for isothermal discs][]{durisen07}. This may be
  possible in the outer parts of protoplanetary discs due to heating
  by irradiation \citep{rafikov07,rice11}.

  % Numerical simulations performed by
  % \cite{rice11} indicate external irradiation can stablise
  % the disc against fragmentation.

  %It is important to note, however,
  %that these studies do not consider the possible effect of global
  %structures present in our numerical simulations.

  %chiang and goldreich
  The background torque may therefore exist in the outer
  parts of protoplanetary discs that are strongly irradiated, such
  that the disc temperature is set externally with a non-zero radial
  gradient.  Of course, if external irradiation sets a strictly
  isothermal outer disc, then the background torque vanishes.

%where does the structure come from? (mass built up, planet gap)

  \subsubsection{Radial disc structure}
  In our numerical simulations the 
  $m=1$ spiral is confined between two $Q$-barriers, where real solutions to the local
  dispersion relation is possible (Eq. \ref{wavenumber}). The
  existence of such a cavity is a result of the adopted disc
  structure. Thus, in our disc models the main role of disc structure
  and self-gravity is to 
  allow a local $m=1$ mode to be set up, which is then 
  destabilised by the background torque density.

  In order to confine an $m=1$ spiral between two $Q$-barriers, we
  should have $Q^2(1-\nu^2)=1$ at two radii. Assuming 
  Keplerian rotation and a slow pattern speed $\Omega_p\ll\Omega$, this amounts to
  \begin{align}\label{qb_cond}
    \left(\frac{R_{Qb}}{R_c}\right)^{-3/2} = 2Q^2(R_{Qb}). 
  \end{align}
  Then it may be possible to have two $Q$-barriers when the $Q$ profile
  of the disc rises more rapidly (decays more slowly) than $R^{-3/2}$
  for decreasing (increasing) $R$.

  %in fall/mass loading? 
  The surface density bump in our initial condition might represent,
  for example, mass built-up in a 
`dead zone' in a protoplanetary
  disc, where the magneto-rotational 
  instability is ineffective for angular momentum transport
  \citep{gammie96,turner08,landry13} and therefore has a reduced mass
  accretion rate. The dead zone may become 
  self-gravitating with sufficient mass built-up
  \citep{armitage01,martin12,martin12b,zhu09,zhu10,zhu10b,bae13,bae14}. 
  However, the thermodynamic conditions in a dead zone 
  %internal heating and cooling can balance such that a global radial
  %temperature gradient is maintained - then background torque may
  %exist.

  %particularly simple functional representation of the bump. a
  %realistic bump may be different

  Another possibility is a gap opened by a giant planet. In that case $Q$
  rises rapidly towards the gap centre since it is a region of low
  surface density, which may satisfy Eq. \ref{qb_cond}. Then the inner
  edge of our bump
  function mimics the outer gap edge. The outer gap edge is then a
  possible site for the growth of a low-frequency one-armed spiral
  through the background torque. 
 %or gap opened by a fragment

% The simulations themselves consider an isothermal or locally isothermal equation of state only.  Gammie (2001) and other work since have shown that the evolution of self-gravitating discs depends both on the stability criterion and on the balance between heating and cooling.  If the cooling rate is fast, it is likely to fragment, otherwise it will settle into a quasi-steady, self-gravitating state.

% When considering isothermal simulations, the likely evolution is one of two things.  It's either initially so unstable that the disc quickly fragments, or it's initially stable enough that any initial perturbations simply drives the system towards being even more stable and the instability quickly weakens through rapid mass transfer (see the work by Laughlin & Bodenheimer 1994/1996).  There is no real way to drive a long-lived quasi-steady, non-fragmenting, self-gravitating state because this is only possible if the system can remain quasi-stable though balancing heating and cooling (see, for example, Lodato & Rice 2004).

% In fact, in the conclusions it is suggested that the conditions used in the simulation would require a cooling < 0.1 \Omega^{-1} which, based on the work of Gammie 2001, would quickly lead to fragmentation.  The reason it doesn't in the work considered here is that the isothermal assumption assumes that there is some equivalently fast heating mechanism but, in the work here, this is not provided by the instability itself.  If it were, the simulations would become very quickly unstable.  Also, if one looks at the work of Clarke (2009), Rafikov (2005), and Rice & Armitage (2009) the only region of a realistic disc where such rapid cooling is possible is in the outer disc, but such regions are either very gravitationally unstable and fragment, or quickly stablised by an external heating source.  I guess it is possible that there could be a scenario where this exactly produces the conditions assumed here, but that would seem rather fine-tuned and it would be more likely that an external heating source would simply stabilise the disc (unless the disc were extremely massive and could remain unstable even at reasonably high disc temperatures).

{\bf
  \subsection{Motion of the central star}

  \subsection{Role of Lindblad and co-rotation torques}
  One possible effect of self-gravity is to allow the one-armed spiral  
  to act as an external potential for the exterior disc. This is
  analogous to disc-satellite interaction 
  \citep{goldreich79}, where the embedded satellite exerts a torque on
  the disc at

In Appendix \ref{disc-planet} we estimate the magnitude of this effect
  using basic results from disc-planet theory \citep[see, e.g.][and
  references therein]{papaloizou07}.

One possible effect of self-gravity is to allow the one-armed spiral  
to act as an external potential for the wider disc. This is
analogous to disc-satellite interaction 
\citep{goldreich79}. Here, we estimate the magnitude of this effect
using basic results from disc-planet theory \citep[see, e.g.][and
references therein]{papaloizou07}.

%Furthermore, this
%non-local interaction may be important for the long-term evolution of
%the outer disc, since it gains angular momentum from the

@@

### 2015-01-07 paper/paper.tex
major revision in response to report

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency
  trailing wave and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the wave.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly-stationary
  one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly depend on self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}

>>>>

\maketitle
\begin{abstract} 
  {\bf We describe a new mechanism that leads to the destabilisation of
 non-axisymmetric waves in astrophysical discs with 
  an imposed radial temperature gradient.  We use linear density wave
  theory 
  to show that non-axisymmetric perturbations generally do not conserve
  their angular momentum in the presence of a forced temperature
  gradient. This implies an exchange of angular momentum
  between linear perturbations and the background disc. In particular, 
  when the disturbance is a low-frequency
  trailing wave and the disc temperature decreases outwards, this 
  interaction is unstable and leads to the growth of the wave.  
  We demonstrate this phenomenon through numerical hydrodynamic
  simulations of locally isothermal discs in 2D using the FARGO
  code and in 3D with  the ZEUS-MP and PLUTO codes.  We consider
  radially structured discs with a self-gravitating sub-portion which remains
  stable in the absence of a temperature gradient. However,
  when a temperature gradient is imposed we observe exponential growth of an one-armed  
  spiral mode (azimuthal wavenumber $m=1$) with co-rotation radius 
  outside the bulk of the spiral arm, resulting in a nearly-stationary
  one-armed spiral pattern.  The development of this one-armed 
 spiral does not require the movement of the central star, as found in previous studies. 
 Because destabilisation by the imposed temperature gradient does 
 not explicitly require disc self-gravity, we suggest this mechanism may also
 affect low-frequency one-armed oscillations in non-self-gravitating
 discs. 
}
\end{abstract}
